---
title: "SISBID: Causal Inference with Machine Learning"
subtitle: "R version"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)

set.seed(42)

library(ggplot2)

theme_set(theme_minimal(base_size = 13))
```

# Average Treatment Effect Estimation

For unit \(i\), let

- \(Y_i^{(1)}\) be the potential outcome under treatment;
- \(Y_i^{(0)}\) be the potential outcome under control;
- \(T_i \in \{0,1\}\) be the treatment indicator.

The average treatment effect is

\[
\tau = \mathbb{E}\left[Y_i^{(1)} - Y_i^{(0)}\right].
\]

For each unit, we observe only one potential outcome:

\[
Y_i = T_iY_i^{(1)} + (1-T_i)Y_i^{(0)}.
\]

# Randomized Treatment

We first consider randomized treatment assignment, so that

\[
\left(Y_i^{(1)},Y_i^{(0)}\right) \perp T_i.
\]

```{r randomized-data}
n_randomized <- 500

# Randomly assign treatment
randomized_t <- rbinom(
  n = n_randomized,
  size = 1,
  prob = 0.5
)

# True treatment effect
randomized_tau <- 4

# Observed outcome
randomized_y <- randomized_tau * randomized_t +
  rnorm(n_randomized)

randomized_data <- data.frame(
  unit = seq_len(n_randomized),
  t = factor(randomized_t, levels = c(0, 1)),
  y = randomized_y
)
```

```{r randomized-plot}
ggplot(
  randomized_data,
  aes(x = unit, y = y, color = t)
) +
  geom_point(alpha = 0.75) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  geom_hline(
    yintercept = randomized_tau,
    linetype = "dashed",
    color = "blue"
  ) +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("Control", "Treatment")
  ) +
  labs(
    x = "Unit",
    y = "Outcome",
    color = "Group",
    title = "Outcomes under randomized treatment"
  )
```

Under random assignment,

\[
\mathbb{E}[Y_i \mid T_i=1]
-
\mathbb{E}[Y_i \mid T_i=0]
=
\mathbb{E}[Y_i^{(1)}-Y_i^{(0)}].
\]

Thus, the ATE can be estimated using the difference in sample means.

```{r randomized-difference-means}
randomized_difference_ate <- 
  mean(randomized_y[randomized_t == 1]) -
  mean(randomized_y[randomized_t == 0])

randomized_difference_ate
```

The same estimate is the coefficient of \(T\) in a linear regression with an intercept.

```{r randomized-regression}
randomized_model <- lm(
  randomized_y ~ randomized_t
)

randomized_regression_ate <- 
  coef(randomized_model)[["randomized_t"]]

coef(randomized_model)
summary(randomized_model)
```

# Observed Confounder

Suppose \(X\) affects both treatment assignment and the outcome. If the outcome model is linear, we may estimate the treatment effect using

\[
Y_i = \alpha + \beta X_i + \tau T_i + \varepsilon_i.
\]

```{r confounder-data}
n_confounder <- 500

confounder_x <- runif(
  n_confounder,
  min = -3,
  max = 3
)

# Treatment is determined by X
confounder_t <- as.integer(confounder_x < 0)

confounder_tau <- 4

confounder_y <- confounder_x +
  confounder_tau * confounder_t +
  rnorm(n_confounder)

confounded_data <- data.frame(
  x = confounder_x,
  t = factor(confounder_t, levels = c(0, 1)),
  y = confounder_y
)
```

```{r confounder-plot}
confounder_line_data <- data.frame(
  x = rep(
    seq(-3, 3, length.out = 200),
    2
  ),
  t = factor(
    rep(c(0, 1), each = 200),
    levels = c(0, 1)
  )
)

confounder_line_data$mean_y <- 
  confounder_line_data$x +
  confounder_tau *
    as.numeric(
      as.character(confounder_line_data$t)
    )

ggplot(
  confounded_data,
  aes(x = x, y = y, color = t)
) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = confounder_line_data,
    aes(x = x, y = mean_y, color = t),
    linewidth = 1
  ) +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("Control", "Treatment")
  ) +
  labs(
    x = "Confounder X",
    y = "Outcome Y",
    color = "Group",
    title = "Treatment assignment depends on X"
  )
```

If we fail to adjust for \(X\), the difference in means is biased.

```{r confounder-unadjusted}
confounder_unadjusted_ate <- 
  mean(confounder_y[confounder_t == 1]) -
  mean(confounder_y[confounder_t == 0])

confounder_unadjusted_ate
```

Adjusting for \(X\) using linear regression recovers the treatment effect.

```{r confounder-adjusted}
adjusted_model <- lm(
  confounder_y ~ confounder_t + confounder_x
)

adjusted_ate <- 
  coef(adjusted_model)[["confounder_t"]]

coef(adjusted_model)
summary(adjusted_model)
adjusted_ate
```

## Frisch--Waugh--Lovell Residualization

The coefficient of \(T\) in the regression of \(Y\) on \(T\) and \(X\) can also be obtained by:

1. residualizing \(Y\) with respect to \(X\);
2. residualizing \(T\) with respect to \(X\);
3. regressing the residualized outcome on the residualized treatment.

```{r fwl}
outcome_residual <- resid(
  lm(confounder_y ~ confounder_x)
)

treatment_residual <- resid(
  lm(confounder_t ~ confounder_x)
)

fwl_model <- lm(
  outcome_residual ~ 0 + treatment_residual
)

fwl_ate <- 
  coef(fwl_model)[["treatment_residual"]]

coef(fwl_model)

c(
  regression_adjustment = adjusted_ate,
  FWL_residualization = fwl_ate
)
```

# Propensity Score

The propensity score is

\[
q(X_i)=\Pr(T_i=1\mid X_i).
\]

Inverse probability weighting can identify the ATE when treatment is conditionally independent of the potential outcomes given \(X\), and when overlap holds:

\[
0 < q(X_i) < 1.
\]

## Simulate Data with Overlap

```{r propensity-data}
n_propensity <- 500

propensity_x <- runif(
  n_propensity,
  min = -3,
  max = 3
)

# True propensity score
true_propensity <- plogis(propensity_x)

propensity_t <- rbinom(
  n_propensity,
  size = 1,
  prob = true_propensity
)

propensity_tau <- 4

propensity_y <- propensity_x +
  propensity_tau * propensity_t +
  rnorm(n_propensity)

propensity_data <- data.frame(
  x = propensity_x,
  t = factor(propensity_t, levels = c(0, 1)),
  y = propensity_y,
  true_propensity = true_propensity
)

range(true_propensity)
```

```{r propensity-outcome-plot}
propensity_line_data <- data.frame(
  x = rep(
    seq(-3, 3, length.out = 200),
    2
  ),
  t = factor(
    rep(c(0, 1), each = 200),
    levels = c(0, 1)
  )
)

propensity_line_data$mean_y <- 
  propensity_line_data$x +
  propensity_tau *
    as.numeric(
      as.character(propensity_line_data$t)
    )

ggplot(
  propensity_data,
  aes(x = x, y = y, color = t)
) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = propensity_line_data,
    aes(x = x, y = mean_y, color = t),
    linewidth = 1
  ) +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("Control", "Treatment")
  ) +
  labs(
    x = "Covariate X",
    y = "Outcome Y",
    color = "Group",
    title = paste(
      "Observed outcomes with probabilistic",
      "treatment assignment"
    )
  )
```

## Estimate the Propensity Score

In practice, the true propensity score is unknown. We estimate it by fitting a logistic regression of treatment on \(X\).

```{r propensity-model}
propensity_model <- glm(
  propensity_t ~ propensity_x,
  family = binomial(link = "logit")
)

estimated_propensity <- predict(
  propensity_model,
  type = "response"
)

summary(propensity_model)
range(estimated_propensity)
```

```{r propensity-comparison}
propensity_comparison <- data.frame(
  estimated = estimated_propensity,
  truth = true_propensity
)

ggplot(
  propensity_comparison,
  aes(x = estimated, y = truth)
) +
  geom_point(alpha = 0.7) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  coord_equal() +
  labs(
    x = "Estimated propensity score",
    y = "True propensity score",
    title = "Estimated versus true propensity scores"
  )
```

## Inverse Probability Weighting

The Horvitz--Thompson IPW estimator is

\[
\widehat{\tau}_{\mathrm{IPW}}
=
\frac{1}{n}
\sum_{i=1}^n
\left[
\frac{T_iY_i}{\widehat q(X_i)}
-
\frac{(1-T_i)Y_i}{1-\widehat q(X_i)}
\right].
\]

```{r ipw-estimator}
ipw_contribution <- 
  propensity_t * propensity_y /
    estimated_propensity -
  (1 - propensity_t) * propensity_y /
    (1 - estimated_propensity)

ipw_ate <- mean(ipw_contribution)

ipw_ate
```

A normalized Hájek version can be more stable in finite samples:

```{r hajek-estimator}
treated_mean_ipw <- 
  sum(
    propensity_t * propensity_y /
      estimated_propensity
  ) /
  sum(
    propensity_t / estimated_propensity
  )

control_mean_ipw <- 
  sum(
    (1 - propensity_t) * propensity_y /
      (1 - estimated_propensity)
  ) /
  sum(
    (1 - propensity_t) /
      (1 - estimated_propensity)
  )

hajek_ate <- treated_mean_ipw - control_mean_ipw

hajek_ate
```

# Colliders

A collider is a variable caused by two other variables. In the following simulation,

\[
T \longrightarrow X \longleftarrow Y.
\]

Conditioning on \(X\) opens a noncausal association between \(T\) and \(Y\). Therefore, we should not adjust for \(X\).

## Collider Example with a Positive Treatment Effect

```{r collider-positive-data}
n_collider_positive <- 500

collider_positive_t <- rbinom(
  n_collider_positive,
  size = 1,
  prob = 0.5
)

collider_positive_tau <- 4

collider_positive_y <- 
  collider_positive_tau * collider_positive_t +
  rnorm(n_collider_positive)

# X is a collider affected by both T and Y
collider_positive_x <- 
  collider_positive_t +
  0.5 * collider_positive_y +
  rnorm(n_collider_positive, sd = 0.2)
```

Without conditioning on the collider, the difference in means estimates the ATE.

```{r collider-positive-unadjusted}
collider_positive_unadjusted_ate <- 
  mean(
    collider_positive_y[
      collider_positive_t == 1
    ]
  ) -
  mean(
    collider_positive_y[
      collider_positive_t == 0
    ]
  )

collider_positive_unadjusted_ate
```

Conditioning on the collider can severely bias the estimate.

```{r collider-positive-adjusted}
collider_positive_model <- lm(
  collider_positive_y ~
    collider_positive_t +
    collider_positive_x
)

collider_positive_adjusted_ate <- 
  coef(
    collider_positive_model
  )[["collider_positive_t"]]

coef(collider_positive_model)
summary(collider_positive_model)
```

## Collider Example with No Treatment Effect

Here, the true treatment effect is zero.

```{r collider-zero-data}
n_collider_zero <- 500

collider_zero_t <- rbinom(
  n_collider_zero,
  size = 1,
  prob = 0.5
)

collider_zero_tau <- 0

collider_zero_y <- 
  collider_zero_tau * collider_zero_t +
  rnorm(n_collider_zero)

collider_zero_x <- 
  collider_zero_t +
  0.5 * collider_zero_y +
  rnorm(n_collider_zero, sd = 0.2)
```

The unadjusted difference in means is approximately unbiased for the zero ATE.

```{r collider-zero-unadjusted}
collider_zero_unadjusted_ate <- 
  mean(
    collider_zero_y[collider_zero_t == 1]
  ) -
  mean(
    collider_zero_y[collider_zero_t == 0]
  )

collider_zero_unadjusted_ate
```

Conditioning on the collider creates a spurious treatment association.

```{r collider-zero-adjusted}
collider_zero_model <- lm(
  collider_zero_y ~
    collider_zero_t +
    collider_zero_x
)

collider_zero_adjusted_ate <- 
  coef(collider_zero_model)[["collider_zero_t"]]

coef(collider_zero_model)
summary(collider_zero_model)
```

# Summary of Estimators

```{r estimator-summary}
results <- data.frame(
  example = c(
    "Randomized: difference in means",
    "Randomized: linear regression",
    "Confounded: unadjusted",
    "Confounded: regression adjustment",
    "Confounded: FWL residualization",
    "Propensity score: Horvitz--Thompson IPW",
    "Propensity score: Hájek IPW",
    "Collider, positive effect: unadjusted",
    "Collider, positive effect: adjusted",
    "Collider, zero effect: unadjusted",
    "Collider, zero effect: adjusted"
  ),
  estimate = c(
    randomized_difference_ate,
    randomized_regression_ate,
    confounder_unadjusted_ate,
    adjusted_ate,
    fwl_ate,
    ipw_ate,
    hajek_ate,
    collider_positive_unadjusted_ate,
    collider_positive_adjusted_ate,
    collider_zero_unadjusted_ate,
    collider_zero_adjusted_ate
  )
)

knitr::kable(
  results,
  digits = 3,
  col.names = c(
    "Method",
    "Estimated treatment effect"
  )
)
```
