---
title: "SISBID3: Classification"
subtitle: "R version"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)

set.seed(42)

library(ggplot2)

theme_set(theme_minimal(base_size = 13))
```


## The Stock Market Data

We load the packages needed for this lab and explore the `Smarket` data set and visualize the correlations among variables. Note that if you try `cor(Smarket)` you will get an error stating `Error in cor(Smarket) : 'x' must be numeric`. This is because `Direction` is a *categorical* variable, hence, not *numeric*. So, below, we remove this variable before looking at correlations. 

```{r}
library(ISLR)
library(MASS)
library(glmnet)
library(class)
library(e1071)
library(ROCR)

names(Smarket)
dim(Smarket)
summary(Smarket)
pairs(Smarket)
cor(Smarket[,-9])
attach(Smarket)
plot(Volume)
```

## Logistic Regression

We fit a logistic regression model to predict `Direction` using the five lag variables and `Volume`.

```{r}
glm.fit=glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,data=Smarket,family=binomial)
summary(glm.fit)
coef(glm.fit)
summary(glm.fit)$coef
```

We compute predicted probabilities, convert them into class labels, and produce a confusion matrix along with the overall accuracy.

```{r}
glm.probs=predict(glm.fit,type="response")
glm.probs[1:10]
contrasts(Direction)
glm.pred=rep("Down",1250)
glm.pred[glm.probs>.5]="Up"
table(glm.pred,Direction)
(507+145)/1250
mean(glm.pred==Direction)
```

## Separate the Data into Training and Test

Next, we create a training set consisting of observations before 2005, and a test set (`Smarket.2005`) consisting of the 2005 observations, then refit the logistic regression on the training set only. This type of data split is especially suitable for *time series* data, i.e., data observed over time, as simple cross validation (or split-sample validation) does not make much sense in this context. 

```{r}
train=(Year<2005)
Smarket.2005=Smarket[!train,]
dim(Smarket.2005)
Direction.2005=Direction[!train]
glm.fit=glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,data=Smarket,family=binomial,subset=train)
glm.probs=predict(glm.fit,Smarket.2005,type="response")
glm.pred=rep("Down",252)
glm.pred[glm.probs>.5]="Up"
table(glm.pred,Direction.2005)
mean(glm.pred==Direction.2005)
mean(glm.pred!=Direction.2005)
```

We refit using only `Lag1` and `Lag2`, the variables that seemed to have the most predictive power, and evaluate on the test set. We also predict the probability of a market increase for two hypothetical days.

```{r}
glm.fit=glm(Direction~Lag1+Lag2,data=Smarket,family=binomial,subset=train)
glm.probs=predict(glm.fit,Smarket.2005,type="response")
glm.pred=rep("Down",252)
glm.pred[glm.probs>.5]="Up"
table(glm.pred,Direction.2005)
mean(glm.pred==Direction.2005)
106/(106+76)
predict(glm.fit,newdata=data.frame(Lag1=c(1.2,1.5),Lag2=c(1.1,-0.8)),type="response")
```

## Penalized Logistic Regression

We set up the predictor matrix and response vector for use with `glmnet`.

```{r}
x = as.matrix(Smarket[1:1250, c(2:7)])
y = as.factor(Smarket[1:1250, 9])
```

### Ridge

We fit a ridge-penalized logistic regression over a grid of lambda values, use cross-validation to select lambda, and evaluate the resulting model on the test set.

```{r}
glm.ridge=glmnet(x[train,],y[train],family="binomial",alpha=0,lambda=exp(seq(-5,10,0.01)))
plot(glm.ridge,col=c(1:8), xvar="lambda")
legend('bottomright',colnames(x),lty=1,col=c(1:8),bty='n')
cv.ridge=cv.glmnet(x[train,],y[train],family="binomial",alpha=0,lambda=exp(seq(-50,2,0.1)))
plot(cv.ridge)
lambda.min=cv.ridge$lambda.min
ridge.probs=predict(glm.ridge,s=lambda.min,type="response",newx=x[!train,])
ridge.pred=rep("Down",252)
ridge.pred[ridge.probs>.5]="Up"
table(ridge.pred,Direction.2005)
mean(ridge.pred==Direction.2005)
```

### Lasso

We repeat the process using a lasso penalty instead. Note that we only need to change `alpha=0` in ridge to `alpha=1`. Everything else is the same. 

```{r}
glm.lasso=glmnet(x[train,],y[train],family="binomial",alpha=1)
plot(glm.lasso,col=c(1:8))
legend('bottomleft',colnames(x),lty=1,col=c(1:8),bty='n')
cv.lasso=cv.glmnet(x[train,],y[train],family="binomial",alpha=1)
plot(cv.lasso)
lambda.min=cv.lasso$lambda.min
lasso.probs=predict(glm.lasso,s=lambda.min,type="response",newx=x[!train,])
lasso.pred=rep("Down",252)
lasso.pred[lasso.probs>.5]="Up"
table(lasso.pred,Direction.2005)
mean(lasso.pred==Direction.2005)
predict(glm.lasso,s=lambda.min,type="coefficients",newx=x[!train,])
```

### Other Choices of Lambda

One nice feature of the `glmnet` package is that it gives you a range of models, so you don't have to refit the model every time you want to try a different tuning parameter `lambda` -- you simply use the predict function and get the results for that value of `lambda`.

```{r}
plot(glm.lasso,col=c(1:8),xvar="lambda")
lasso.probs=predict(glm.lasso,s=exp(-5),type="response",newx=x[!train,])
lasso.pred=rep("Down",252)
lasso.pred[lasso.probs>.5]="Up"
table(lasso.pred,Direction.2005)
mean(lasso.pred==Direction.2005)
```

## K-Nearest Neighbors

Instead of the parametric approach used in logistic regression, we can use a fully nonparametric classification model based on K-nearest neighbors (KNN). In R, we can do this by using the `knn()` function. 

Given its nonparametric nature, KNN cannot be fit with a large number of covariates, and is only applicable to small number of covariates, so here we use `Lag1` and `Lag2` and try different values of neighborhood size, `k`.

```{r}
library(class)
train.X=cbind(Lag1,Lag2)[train,]
test.X=cbind(Lag1,Lag2)[!train,]
train.Direction=Direction[train]
set.seed(1)
knn.pred=knn(train.X,test.X,train.Direction,k=1)
table(knn.pred,Direction.2005)
(83+43)/252
knn.pred=knn(train.X,test.X,train.Direction,k=3)
table(knn.pred,Direction.2005)
mean(knn.pred==Direction.2005)
```

## Another Example: Classification with the Heart Data

We load the `Heart` data set, available on GitHub, and explore it.

*Note: `Heart.rda` must be present in the working directory when knitting.*

```{r}
load("Heart.rda")
attach(Heart)

names(Heart)
dim(Heart)
summary(Heart)
plot(trestbps)
```

## Separate the Data into Training and Test

```{r}
set.seed(1)
train=sample(1:nrow(Heart), 197)
trndat=Heart[train,]
tstdat=Heart[-train,]
dim(trndat)
dim(tstdat)
```

## Logistic Regression

We fit a logistic regression model to predict heart disease (`HD`) using all predictors, and evaluate it on the test set.

```{r}
HD.tst=HD[-train]
glm.fit=glm(HD~.,data=Heart,family=binomial,subset=train)
glm.probs=predict(glm.fit,tstdat,type="response")
glm.pred=numeric(100)
glm.pred[glm.probs>.5]=1
table(glm.pred,HD.tst)
mean(glm.pred==HD.tst)
```

## Penalized Logistic Regression

```{r}
n=nrow(Heart)
x=as.matrix(Heart[1:n,1:13])
y=as.factor(Heart[1:n,14])
```

### Ridge 

```{r}
glm.ridge=glmnet(x[train,],y[train],family="binomial",alpha=0)
plot(glm.ridge,col=c(1:8), xvar="lambda")
legend('topright',colnames(x),lty=1,col=c(1:8),bty='n')
cv.ridge=cv.glmnet(x[train,],y[train],family="binomial",alpha=0)
plot(cv.ridge)
lambda.min=cv.ridge$lambda.min
ridge.probs=predict(glm.ridge,s=lambda.min,type="response",newx=x[-train,])
ridge.pred=rep(0,100)
ridge.pred[ridge.probs>.5]=1
table(ridge.pred,HD.tst)
mean(ridge.pred==HD.tst)
```

### Lasso

```{r}
glm.lasso=glmnet(x[train,],y[train],family="binomial",alpha=1)
plot(glm.lasso,col=c(1:8), xvar="lambda")
legend('bottomright',colnames(x),lty=1,col=c(1:8),bty='n')
cv.lasso=cv.glmnet(x[train,],y[train],family="binomial",alpha=1)
plot(cv.lasso)
lambda.min=cv.lasso$lambda.min
lasso.probs=predict(glm.lasso,s=lambda.min,type="response",newx=x[-train,])
lasso.pred=rep(0,100)
lasso.pred[ridge.probs>.5]=1
table(lasso.pred,HD.tst)
mean(lasso.pred==HD.tst)
```

## K-Nearest Neighbors

```{r}
library(class)
train.X=cbind(ca,thal)[train,1:2]
test.X=cbind(ca,thal)[-train,1:2]
HD.trn=HD[train]
set.seed(1)
knn.pred=knn(train.X,test.X,as.factor(HD.trn),k=1)
table(knn.pred,HD.tst)
mean(knn.pred==HD.tst)
knn.pred=knn(train.X,test.X,HD.trn,k=10)
table(knn.pred,HD.tst)
mean(knn.pred==HD.tst)
```

## SVM

We fit a support vector machine with a radial kernel to the Heart data, then use `tune.svm()` to select the cost and gamma parameters via cross-validation.

```{r}
dat=Heart
dat$HD=as.factor(HD)
svmfit=svm(HD~.,data=dat[train,],kernel="radial",gamma=1,cost=1)
svm.pred=predict(svmfit,newdata=dat[-train,-14])
table(true=dat[-train,14], pred=svm.pred)
mean(dat[-train,14]==svm.pred)

set.seed(1)
tune.out=tune.svm(x=dat[train,-14],y=dat[train,14],kernel="radial",
                  cost=c(0.1,1,10,100,1000),gamma=c(0.5,1,2,3,4))
summary(tune.out)
svmfit=svm(HD~.,data=dat[train,],kernel="radial",gamma=tune.out$best.parameters$gamma,cost=tune.out$best.parameters$cost)
svm.pred=predict(svmfit,newdata=dat[-train,-14])
table(true=dat[-train,14], pred=svm.pred)
mean(dat[-train,14]==svm.pred)
```

## ROC Curves

We define a helper function to plot ROC curves from fitted decision values, and compare an "optimal" versus a more flexible SVM fit on the training and test data.

*Note: this section references a `y` column and `svmfit.opt`/`svmfit.flex` objects that are not defined elsewhere in this script for the `Heart`/`dat` data (`dat` has no `y` column, only `HD`). This looks like it may have been carried over from a different example and will likely error as written; you may want to adapt the formula (`HD~.`) and variable name (`HD` instead of `y`) to match this data set.*

```{r}
rocplot=function(pred, truth, ...){
  predob = prediction(pred, truth)
  perf = performance(predob, "tpr", "fpr")
  plot(perf,...)}

svmfit.opt=svm(HD~., data=dat[train,], kernel="radial", gamma=2, cost=1, decision.values=T)
fitted=attributes(predict(svmfit.opt,dat[train,],decision.values=TRUE))$decision.values
par(mfrow=c(1,2))
rocplot(fitted,dat[train,"HD"],main="Training Data")
svmfit.flex=svm(HD~., data=dat[train,], kernel="radial",gamma=50, cost=1, decision.values=T)
fitted=attributes(predict(svmfit.flex,dat[train,],decision.values=T))$decision.values
rocplot(fitted,dat[train,"HD"],add=T,col="red")
legend('bottomright',c('optimal model','flexible model'),lty=1,col=c(1,2),bty='n')
fitted=attributes(predict(svmfit.opt,dat[-train,],decision.values=T))$decision.values
rocplot(fitted,dat[-train,"HD"],main="Test Data")
fitted=attributes(predict(svmfit.flex,dat[-train,],decision.values=T))$decision.values
rocplot(fitted,dat[-train,"HD"],add=T,col="red")
```
