---
title: "SISBID3: Cross Validation"
subtitle: "R version"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)

set.seed(42)

library(ggplot2)

theme_set(theme_minimal(base_size = 13))
```

## Cross-Validation

We begin by loading the `ISLR` package and examining the dimensions of the `Auto` data set.

```{r}
library(ISLR)
ncol(Auto)
nrow(Auto)
```
## Plot data
Often times, it's good to practice to plot the data you are trying to analyze.This gives you a sense of whether there are outlier, interesting patterns, etc. These can motivate the model you end up fitting.

```{r}
set.seed(1)
train=sample(392,196)
plot(mpg ~ horsepower,
     data = Auto,
     xlab = "Horsepower",
     ylab = "MPG",
     main = "MPG versus Horsepower",
     pch = 19)
```

## The Validation Set Approach

We use the validation set approach to estimate the test error rates from fitting various linear models on the `Auto` data set. We first split the observations into a training set and a validation set.

```{r}
set.seed(1)
train=sample(392,196)
```

We then fit a linear regression using only the training observations.

```{r}
lm.fit=lm(mpg~horsepower,data=Auto,subset=train)
attach(Auto)
```

We use `predict()` to estimate the response for all 392 observations, and we calculate the mean squared error of the 196 observations in the validation set.

```{r}
mean((mpg-predict(lm.fit,Auto))[-train]^2)
```

We can use `poly()` to estimate the test error for the quadratic and cubic regressions.

```{r}
lm.fit2=lm(mpg~poly(horsepower,2),data=Auto,subset=train)
mean((mpg-predict(lm.fit2,Auto))[-train]^2)
lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)
mean((mpg-predict(lm.fit3,Auto))[-train]^2)
```

If we choose a different training set instead, then we will obtain somewhat different errors on the validation set.

```{r}
set.seed(2)
train=sample(392,196)
lm.fit=lm(mpg~horsepower,subset=train)
mean((mpg-predict(lm.fit,Auto))[-train]^2)
lm.fit2=lm(mpg~poly(horsepower,2),data=Auto,subset=train)
mean((mpg-predict(lm.fit2,Auto))[-train]^2)
lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)
mean((mpg-predict(lm.fit3,Auto))[-train]^2)
```

## Leave-One-Out Cross-Validation

The LOOCV estimate can be automatically computed for any generalized linear model using the `glm()` and `cv.glm()` functions. We first show that `glm()` yields identical results to `lm()` when no `family` argument is passed.

```{r}
glm.fit=glm(mpg~horsepower,data=Auto)
coef(glm.fit)
lm.fit=lm(mpg~horsepower,data=Auto)
coef(lm.fit)
```

We now use the `glm()` function together with `cv.glm()` from the `boot` package to perform cross-validation.

```{r}
library(boot)
glm.fit=glm(mpg~horsepower,data=Auto)
cv.err=cv.glm(Auto,glm.fit)
cv.err$delta
```

We repeat this procedure for increasingly complex polynomial fits, iteratively fitting the polynomial regression for polynomials of order 1 through 5, computing the associated cross-validation error, and plotting the results.

```{r}
cv.error=rep(0,5)
for (i in 1:5){
 glm.fit=glm(mpg~poly(horsepower,i),data=Auto)
 cv.error[i]=cv.glm(Auto,glm.fit)$delta[1]
 }
plot(cv.error)
```

## k-Fold Cross-Validation

The `cv.glm()` function can also be used to implement k-fold CV. We set a seed and initialize a vector in which we will store the CV errors corresponding to the polynomial fits of orders one to ten.

```{r}
set.seed(17)
cv.error.10=rep(0,10)
for (i in 1:10){
 glm.fit=glm(mpg~poly(horsepower,i),data=Auto)
 cv.error.10[i]=cv.glm(Auto,glm.fit,K=10)$delta[1]
 }
plot(cv.error.10)
```

Finally, we summarize the quadratic model.

```{r}
glm.fit2=glm(mpg~poly(horsepower,2),data=Auto)
summary(glm.fit2)
```

## Model Complexity and Variance

More flexible models typically have higher variance: small changes in the training data can produce larger changes in the fitted model. To illustrate this, we fit a linear model and a tenth-degree polynomial using two different training sets.

We first standardize horsepower to improve the numerical stability of the polynomial regression.

```{r}
horsepower.mean = mean(Auto$horsepower)
horsepower.sd = sd(Auto$horsepower)

Auto$horsepower.std =
  (Auto$horsepower - horsepower.mean) / horsepower.sd
```

We construct two training sets, each containing half of the observations.

```{r}
set.seed(1)
train.1 = sample(seq_len(nrow(Auto)), nrow(Auto) / 2)

set.seed(2)
train.2 = sample(seq_len(nrow(Auto)), nrow(Auto) / 2)
```

We fit a linear regression and a tenth-degree polynomial regression to each training set.

```{r}
linear.fit.1 = lm(
  mpg ~ horsepower.std,
  data = Auto,
  subset = train.1
)

linear.fit.2 = lm(
  mpg ~ horsepower.std,
  data = Auto,
  subset = train.2
)

poly.fit.1 = lm(
  mpg ~ poly(horsepower.std, 10, raw = TRUE),
  data = Auto,
  subset = train.1
)

poly.fit.2 = lm(
  mpg ~ poly(horsepower.std, 10, raw = TRUE),
  data = Auto,
  subset = train.2
)
```

We first compare the coefficients of the two linear models.

```{r}
linear.coefs = data.frame(
  term = c("Intercept", "Horsepower"),
  training_set_1 = coef(linear.fit.1),
  training_set_2 = coef(linear.fit.2)
)

linear.coefs$absolute_difference =
  abs(linear.coefs$training_set_1 -
        linear.coefs$training_set_2)

knitr::kable(
  linear.coefs,
  digits = 3,
  row.names = FALSE
)
```

We can similarly compare the coefficients of the tenth-degree polynomial models.

```{r}
poly.coefs = data.frame(
  term = c("Intercept", paste0("Degree ", 1:10)),
  training_set_1 = coef(poly.fit.1),
  training_set_2 = coef(poly.fit.2)
)

poly.coefs$absolute_difference =
  abs(poly.coefs$training_set_1 -
        poly.coefs$training_set_2)

knitr::kable(
  poly.coefs,
  digits = 3,
  row.names = FALSE
)
```

Because polynomial coefficients can be difficult to interpret directly, we also compare the fitted regression curves.

```{r}
horsepower.grid = seq(
  min(Auto$horsepower),
  max(Auto$horsepower),
  length.out = 300
)

prediction.grid = data.frame(
  horsepower.std =
    (horsepower.grid - horsepower.mean) / horsepower.sd
)

prediction.curves = rbind(
  data.frame(
    horsepower = horsepower.grid,
    predicted_mpg = predict(linear.fit.1, prediction.grid),
    model = "Linear model",
    training_set = "Training set 1"
  ),
  data.frame(
    horsepower = horsepower.grid,
    predicted_mpg = predict(linear.fit.2, prediction.grid),
    model = "Linear model",
    training_set = "Training set 2"
  ),
  data.frame(
    horsepower = horsepower.grid,
    predicted_mpg = predict(poly.fit.1, prediction.grid),
    model = "Degree-10 polynomial",
    training_set = "Training set 1"
  ),
  data.frame(
    horsepower = horsepower.grid,
    predicted_mpg = predict(poly.fit.2, prediction.grid),
    model = "Degree-10 polynomial",
    training_set = "Training set 2"
  )
)
```

```{r}
ggplot(Auto, aes(x = horsepower, y = mpg)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = prediction.curves,
    aes(
      x = horsepower,
      y = predicted_mpg,
      linetype = training_set
    ),
    linewidth = 1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ model) +
  labs(
    title = "Sensitivity to the Choice of Training Data",
    x = "Horsepower",
    y = "Predicted MPG",
    linetype = NULL
  )
```

Finally, we calculate the root mean squared difference between the fitted curves.

```{r}
linear.pred.1 = predict(linear.fit.1, prediction.grid)
linear.pred.2 = predict(linear.fit.2, prediction.grid)

poly.pred.1 = predict(poly.fit.1, prediction.grid)
poly.pred.2 = predict(poly.fit.2, prediction.grid)

fit.variability = data.frame(
  model = c(
    "Linear model",
    "Degree-10 polynomial"
  ),
  RMSE_between_fitted_curves = c(
    sqrt(mean((linear.pred.1 - linear.pred.2)^2)),
    sqrt(mean((poly.pred.1 - poly.pred.2)^2))
  )
)

knitr::kable(
  fit.variability,
  digits = 3,
  row.names = FALSE
)
```

The tenth-degree polynomial generally changes more substantially across the two training sets. This illustrates the higher variance of flexible models: their fitted predictions are more sensitive to which observations are included in the training data.
